Simple position restraint schemes for alchemical binding free energy calculations

### 2022-01-29 manuscript/position-restraints.tex

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[%
 aip,
%jmp,%
%bmf,%
%sd,%
rsi,%
 amsmath,amssymb,
%preprint,%
 reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

\usepackage{xcolor}

\usepackage{todonotes}

@@

Currently, the most popular restraint scheme is a method by Boresch et al \cite{boresch2003absolute} that utilizes a set of orthogonal harmonic restraints.  The advantage of this scheme is that it provides an exact analytical expression for the free energy of restraint for a ligand in solution.    There are several disadvantages to this scheme, however.  One disadvantage is the practical difficulty of implementing different kinds of restraints.  The method uses one distance restraint, two angle restraints, and three dihedral restraints, requiring the choice of six different force constant parameters.  Furthermore, great care is required to judiciously choose the degrees of freedom on which to apply these constraints.  This difficulty has recently been emphasized by Proccaci et al. \cite{procacci2021ns} who show that large biases in free energies estimated using a non-equilibrium work (NEW) method can be introduced when the Boresch-restrained ligand pose does not match well with the actual pose, which is often unknown before performing the calculation.

@@

In traditional free energy perturbation approaches, a number of simulations are performed in different thermodynamic ensembles indexed by $\lambda$, corresponding to (reduced) potential energy functions $u_{\lambda}(x) = \beta U_{\lambda}(x)$, where $\beta = 1/kT$.  For instance, to calculate a molecule's free energy of solvation, the $\lambda$ parameter might interpolate between 0 and 1 for potential energy functions $u_{\lambda}(x) = u_{\text{bonded}}(x) + \lambda u_{\text{nonbonded}}(x)$, for a simulation of the molecule in solution.  Due to the statistical inefficiencies of exponential averaging, it is very difficult to directly estimate the free energy $f = -\ln Z_{\lambda=1}/Z_{\lambda=0}$, where $Z_{\lambda} = \int e^{-u_{\lambda}(x)} dx$ is the partition function.  Instead,  the calculation is performed at many intermediate $\lambda$ values (e.g. $\lambda_i = 0, 0.05, 0.1, .... 1$), with the energies computed from these simulations used as input to one of several multi-ensemble free energy estimators.\cite{shirts2008statistically, TI}

@@

% \nocite{*}
\bibliography{main}% Produces the bibliography via BibTeX.

@@

Thermodynamic ensembles biased in this way also have equal Monte Carlo acceptance probabilities $p_{ij}$ and $p_{ji}$.  Assuming a simple Metropolis criterion for accepting a Monte Carlo move, the ensemble-average acceptance of a move $i \rightarrow j$ is:
\begin{align}
    p_{ij} &= \frac{ \int \min [ 1, e^{-\Delta u_{ij}(x) + \Delta f_{ij} } ] e^{-u_i(x)} dx} {Z_1} \\
    &= \frac{\int_{\Delta u_{ij} < \Delta f_{ij} } (1) e^{-u_i(x)} dx}{Z_1} + \frac{\int_{\Delta u_{ij} \geq \Delta f_{ij} } e^{-u_j(x)} e^{\Delta f_{ij}} dx}{Z_1} \\
    &=  \frac{Z_1}{Z_2} \frac{\int_{\Delta u_{ji} > \Delta f_{ji} } e^{-u_i(x)} e^{\Delta f_{ji}} dx}{Z_1} + \frac{Z_1}{Z_2} \frac{\int_{\Delta u_{ij} \geq \tilde{f}_{ij} } e^{-u_j(x)} dx}{Z_1} \\
    &=  \frac{\int_{\Delta u_{ji} > \Delta f_{ji} } e^{-\Delta u_{ji}(x)} e^{-u_j(x)} dx}{Z_2} + \frac{\int_{\Delta u_{ji} \leq \Delta \tilde{f}_{ji} } (1) e^{-u_j(x)} dx}{Z_2} \\
    &= \frac{ \int \min [ 1, e^{-\Delta u_{ji}(x) + \Delta f_{ji} } ] e^{-u_j(x)} dx} {Z_2} = p_{ji}. \\
\end{align}

\newpage

>>>>

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[%
 aip,
 rsi,
 jmp,%
 amsmath,amssymb,
%preprint,%
 reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

\usepackage{graphicx}   % Include figure files
\usepackage{dcolumn  }% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

\usepackage{xcolor}
\setlength{\marginparwidth}{2cm} 
\usepackage{todonotes}

@@

Currently, the most popular restraint scheme is a method by Boresch et al. \cite{boresch2003absolute} that utilizes a set of orthogonal harmonic restraints.  The advantage of this scheme is that it provides an exact analytical expression for the free energy of restraint for a ligand in solution.    There are several disadvantages to this scheme, however.  One disadvantage is the practical difficulty of implementing different kinds of restraints.  The method uses one distance restraint, two angle restraints, and three dihedral restraints, requiring the choice of six different force constant parameters.  Furthermore, great care is required to judiciously choose the degrees of freedom on which to apply these constraints.  This difficulty has recently been emphasized by Proccaci et al. \cite{procacci2021ns} who show that large biases in free energies estimated using a non-equilibrium work (NEW) method can be introduced when the Boresch-restrained ligand pose does not match well with the actual pose, which is often unknown before performing the calculation.

@@

In traditional free energy perturbation approaches, a number of simulations are performed in different thermodynamic ensembles indexed by $\lambda$, corresponding to (reduced) potential energy functions $u_{\lambda}(x) = \beta U_{\lambda}(x)$, where $\beta = 1/kT$.  For instance, to calculate a molecule's free energy of solvation, the $\lambda$ parameter might interpolate between 0 and 1 for potential energy functions $u_{\lambda}(x) = u_{\text{bonded}}(x) + \lambda u_{\text{nonbonded}}(x)$, for a simulation of the molecule in solution.  Due to the statistical inefficiencies of exponential averaging, it is very difficult to directly estimate the free energy $f = -\ln Z_{\lambda=1}/Z_{\lambda=0}$, where $Z_{\lambda} = \int e^{-u_{\lambda}(x)} dx$ is the partition function.  Instead,  the calculation is performed at many intermediate $\lambda$ values (e.g. $\lambda_i = 0, 0.05, 0.1, .... 1$), with the energies computed from these simulations used as input to one of several multi-ensemble free energy estimators.\cite{shirts2008statistically,TI}

@@

\section*{References}
% \nocite{*}

\bibliography{main}% Produces the bibliography via BibTeX.
%\bibliography{aipsamp}% Produces the bibliography via BibTeX.

@@

Thermodynamic ensembles biased in this way also have equal Monte Carlo acceptance probabilities $p_{ij}$ and $p_{ji}$.  Assuming a simple Metropolis criterion for accepting a Monte Carlo move, the ensemble-average acceptance of a move $i \rightarrow j$ is:
\begin{align}
    p_{ij} &= \frac{ \int \min [ 1, e^{-\Delta u_{ij}(x) + \Delta f_{ij} } ] e^{-u_i(x)} dx} {Z_1} \\
    &= \frac{\int_{\Delta u_{ij} < \Delta f_{ij} } (1) e^{-u_i(x)} dx}{Z_1} + \frac{\int_{\Delta u_{ij} \geq \Delta f_{ij} } e^{-u_j(x)} e^{\Delta f_{ij}} dx}{Z_1} \\
    &=  \frac{Z_1}{Z_2} \frac{\int_{\Delta u_{ji} > \Delta f_{ji} } e^{-u_i(x)} e^{\Delta f_{ji}} dx}{Z_1} + \frac{Z_1}{Z_2} \frac{\int_{\Delta u_{ij} \geq \tilde{f}_{ij} } e^{-u_j(x)} dx}{Z_1} \\
    &=  \frac{\int_{\Delta u_{ji} > \Delta f_{ji} } e^{-\Delta u_{ji}(x)} e^{-u_j(x)} dx}{Z_2} + \frac{\int_{\Delta u_{ji} \leq \Delta \tilde{f}_{ji} } (1) e^{-u_j(x)} dx}{Z_2} \\
    &= \frac{ \int \min [ 1, e^{-\Delta u_{ji}(x) + \Delta f_{ji} } ] e^{-u_j(x)} dx} {Z_2} = p_{ji}. 
\end{align}

% \newpage
